### 2020-02-29 doc/icfp20/counterpoint.tex
wip
better figure for motion
separate figures of 4 motions and 2 bad patterns
tweak counterpoint section

\newcommand{\Motion}{
  \begin{figure}
    \includegraphics[width=6cm]{fig/motion.png}
    \hspace{1cm}
    \includegraphics[width=3cm]{fig/badmotion.png} \\
    \begin{flushleft}
    \begin{small}
      \hspace{1.95cm} contrary \ \ \ \ parallel \ \ \ \ \ \ \ similar \ \ \ \ \ \ oblique
      \hspace{2.2cm} 5th \hspace{9.5mm} 8th
    \end{small}
    \end{flushleft}
    \caption{Left: Four Kinds of Motion, Right: Invalid Motion}
    \label{fig:motion}
  \end{figure}
}

@@

Counterpoint is a technique for combining multiple lines of melodies.
Composing counterpoint is like arranging a song for choir:
we start with a \emph{cantus firmus}, which serves as the base melody,
and compose counterpoint lines above or below the cantus firmus.
When doing this, we must make sure that the whole music sounds
harmonically pleasing, and that the individual melodic lines are
distinguishable to the listener.

In this section, we present an implementation of species counterpoint,
based on the formulation given by Fux~\citep{fux-cp}.
The idea is to represent ``good'' counterpoint  as a dependent record,
whose fields encode the musical content as well as proofs that the
counterpoint follows certain rules.
For space reason, we only describe two variants of species counterpoint;
other variants can be formalized in an analogous way.

@@

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which is required to start with a tonic and consist only of whole notes.
Figure~\ref{fig:fs} shows an example of first species counterpoint.
The lower line is the cantus firmus, which we excerpt from a famous
German song called ``Froschgesang'' (Frog's song).
The upper line is the counterpoint composed by the second author.

@@

The beginning of the music should express perfection.
As we saw in Section~\ref{sec:music}, there are three intervals that
are classified as perfect: the 1st (more commonly called the
\emph{unison}), the 5th, and the 8th.
The first interval of the music must then be one of these intervals.
In our formalization, we implement this rule as the
\texttt{checkBeginning} function, which reports an error
\texttt{not158 i} when the first interval \texttt{i} is an invalid one.
Since Agda does not have exceptions, we turn the error into an option
value by wrapping it around the \texttt{just} constructor.

@@

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As can be seen from Figure~\ref{fig:motion} left, there are four
kinds of motion: \emph{contrary} (two lines go in different directions),
\emph{parallel} (two lines go in the same direction by the same
distance), \emph{similar} (two lines go in the same direction),
and \emph{oblique} (one line plays the same note).
There is a consensus that approaching a perfect interval by parallel
or similar motion (as in Figure~\ref{fig:motion} right) destroys the
independence of lines.
To rule out this, we define the \texttt{checkMotion} function, which
lists all occurrences of such invalid motion.

@@

checkMotion : List PitchInterval \(\rightarrow\) List MotionError
checkMotion = mapMaybe (uncurry motionCheck) \(\circ\) pairs
-- where pairs returns a list of all adjacent pairs in a given list
\end{alltt}

@@

The ending of the music should express relaxation.
Among different intervals, the unison and the 8th are considered
most stable, hence the last interval of the music must be either of
these intervals.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution.
We encode these rules as the \texttt{checkEnding} function.
Note that it returns the \texttt{tooShort} error when the music does
not have enough number of bars to form a valid ending.

>>>>

\newcommand{\Motion}{
  \begin{figure}
    \includegraphics[width=7cm]{fig/motion1.png}
    \hspace{8mm}
    \includegraphics[width=3.35cm]{fig/motion2.png}
    \begin{flushleft}
    \begin{small}
      \hspace{1.55cm} contrary \ \ \ \ \ \ \ \ parallel \ \ \ \ \ \ \
      \ \ \,similar \ \ \ \ \ \ \ \ \ oblique
      \hspace{1.3cm} paralell-5th \ \ \ similar-8th
    \end{small}
    \end{flushleft}
    \caption{Four Kinds of Motion (left) and Invalid Patterns (right)}
    \label{fig:motion}
  \end{figure}
}

@@

TODO: contrast old and new methods

Counterpoint is a technique for combining multiple lines of melodies.
Composing counterpoint is like arranging a song for choir:
we start with a \emph{cantus firmus}, which serves as the base melody,
and compose a counterpoint line above or below the cantus firmus.
When doing this, we must make sure that the whole music sounds
harmonically pleasing, and that the individual melodic lines are
distinguishable to the listener.

In this section, we present an implementation of species counterpoint,
based on the formulation given by \citet{fux-cp}.
The idea is to represent ``good'' counterpoint  as a dependent record,
whose fields encode the musical content as well as proofs that the
counterpoint follows certain rules.
For space reason, we only describe two variants of species counterpoint;
other variants can be formalized in an analogous way.

@@

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which is required to start with a tonic (the first note of a scale;
e.g., C in the case of C major) and consist only of whole notes.
Figure~\ref{fig:fs} shows an example of first species counterpoint.
The lower line is the cantus firmus, which we excerpt from a German
song called \emph{Froschgesang} (Frog's song).
The upper line is the counterpoint composed by the second author.

@@

The beginning of the music should express perfection.
As we saw in Section~\ref{sec:music}, there are three intervals that
are classified as perfect: the 1st (usually called the \emph{unison}),
t5th, and 8th.
The first interval of the music must then be one of these intervals.
In our formalization, we implement this rule as the
\texttt{checkBeginning} function, which reports an error
\texttt{not158 i} when the first interval \texttt{i} is an invalid one.
Since Agda does not have exceptions, we turn the error into an option
value by wrapping it around the \texttt{just} constructor.

@@

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As Figure~\ref{fig:motion} left shows, there are four kinds of motion:
\emph{contrary} (two lines go in different directions),
\emph{parallel} (two lines go in the same direction by the same
distance), \emph{similar} (two lines go in the same direction),
and \emph{oblique} (one line plays the same note).
It is known that approaching a perfect interval by parallel or similar
motion (as in Figure~\ref{fig:motion} right) destroys the independence
of lines.
To rule out such motion, we define the \texttt{checkMotion}
function, which checks all pairs of two adjacent intervals and
returns a list of erroneous patterns.

@@

checkMotion : List PitchInterval \(\rightarrow\) List MotionError
checkMotion = mapMaybe (uncurry motionCheck) \(\circ\) pairs
-- where pairs l returns a list of pairs consisting of two adjacent elements of l
\end{alltt}

@@

The ending of the music should express relaxation.
Among different intervals, the unison and 8th are considered
most stable, hence the last interval of the music must be either of
these intervals.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution.
This in turn suggests that a valid ending requires the middle bars
to be non-empty.
We encode these rules as the \texttt{checkEnding} function, which,
upon finding an invalid ending, tells us which rules are not satisfied.
